---
title: "Homework 1"
output: html_notebook
student: Rodolfo Lerma
---

DELIVERABLE 1

As tropical cyclones make their way over the Atlantic and Pacific oceans during hurricane season, having an accurate forecast ahead of time is crucial in providing decision-makers the opportunity to undertake life-saving actions. Throughout your Masters program, you will learn various predictive tools that are useful in a wide variety of situations. In this deliverable, we focus on descriptive analytics pertaining to past tropical cyclones.

A tropical cyclone is a storm characterized by strong winds and heavy rain. Tropical cyclones originate from the oceans in tropical areas and grow bigger as they move. Some tropical cyclones make landfall (i.e., they touch land) and they are very closely monitored as they can be life-threatening and destructive.

A tropical cyclone is classified into one of 7 categories based on its wind speed (see table below):

- A category 0 tropical cyclone is also called a tropical depression 
- A category 1 tropical cyclone is also called a tropical storm 
- A tropical cyclone in categories 2-6 is called a hurricane

```{r, message = FALSE}
library(skimr)
library(tidyverse)
```

```{r, message = FALSE}
df <- read_csv("tropical_cyclones.csv")
```

```{r}
head(df)
```

```{r}
glimpse(df)
```

1.  Which month of the year has the highest number of tropical cyclones making landfall?

```{r}
df %>% count(month)
```

We can see that the month with the highest tropical cyclones making landfall is the month of SEPTEMBER with 83 times tropical cyclones over the 40 years period represented here.

2.  Compare the tropical cyclone activity across the North Atlantic and the Eastern Pacific basins:

i.  Which of the two basins has, on average, the more severe tropical cyclones as measured by the maximum sustained wind speed at landfall?

```{r}
df_grouped <-  group_by(df, basin)
summary_avg_speed <- summarize(df_grouped, avg_wing_speed = mean(wind_speed_landfall), sd_wing_speed = sd(wind_speed_landfall))
summary_avg_speed
```
It seems that the North Atlantic in AVG gets slightly stronger winds, with a similar STD DEV in both cases.

ii. Which of the two basins has the higher proportion of hurricanes among all tropical cyclones originating in that basin (a hurricane is indicated by 1 in the hurricane_landfall column)?

```{r}
summary_avg_basin_hurr <- summarize(df_grouped, sum_hurricanes = sum(hurricane_landfall))
summary_avg_basin_hurr
```
We can see based on the data above that the North Atlantic has higher proportion of Tropical Cyclones that end up in Hurricanes (65 compared to 31 from the Eastern Pacific)

3.  Assume that the maximum sustained wind speed at landfall of a tropical cyclone follows a normal distribution:

i.  Estimate and report the mean and standard deviation of that distribution across all tropical cyclones in the data set.

```{r}
avg <- mean(df$wind_speed_landfall)
std_dev <- sd(df$wind_speed_landfall)
cat('avg:', avg, 'std_dev:', std_dev)
```
We can see that the mean and std deviation is 56.52 knots with a variation (standard deviation) of 25.66 knots.

```{r}
hist(df$wind_speed_landfall)
```

ii. Using the mean and standard deviation you computed in (i), calculate the probability that the maximum sustained wind speed at landfall is less than or equal to 34 knots (that is, the cyclone is a tropical depression)? How does this compare with the actual proportion of tropical depressions in the dataset?

```{r}
#Base on a Normal distribution assumption
proba_34_distribution = 100*pnorm(34, mean = avg, sd = std_dev, lower.tail = TRUE)
proba_34_distribution
```
Based on the mean and std and under the assumption of normal distribution the probability of the wind speed being less of equal to 34 knots is 19%.

```{r}
smaller_df = df[df$wind_speed_landfall <= 34.0,]
proba_34_data = 100*nrow(smaller_df)/nrow(df)
proba_34_data
```
Based on the data the probability that the wind speed at landfall is less or equal to 34 is 20%. Then the difference is 1% between the calculated probability and the one given by the data.

4.  Your dataset contains a damage column that represents the total damage (in \$billions) caused by a tropical cyclone when it made landfall. We provide below an R command and output that calculates the average of that column by basin. One interpretation of this output is that cyclones from the North Atlantic basin are approximately ten times stronger than cyclones from the Eastern Pacific basin. What are some other reasons that possibly explain the difference in average damage observed in this output?

```{r}
avg_cost = summarize(group_by(df, basin), avg_damage = mean(damage, na.rm = TRUE), sd_damage = sd(damage, na.rm = TRUE), avg_speed = mean(wind_speed_landfall))
avg_cost
```

```{r}
pacific_df = df[df$basin == 'Eastern Pacific',]
hist(pacific_df$damage,breaks = seq(0,6,0.1))
```
Based on the plot above we can notice that we only have 2 data points for Eastern Pacific. While for the North Atlantic we have around 40 data points available.

```{r}
atlantic_df = df[df$basin == 'North Atlantic',]
hist(atlantic_df$damage, breaks = seq(0,150,2))
```

```{r}
null_values = summarize(group_by(df, basin), missing_val = sum(is.na(damage)), number_rows = n(), ratio_missing_rows = sum(is.na(damage)/n()))
null_values
```
We can see that the ratio of missing data is 97% for the Pacific Tropical Cyclone while for the Atlantic is 78%.

```{r}
plot(df$wind_speed_landfall, df$damage,
     pch = 19,
     col = factor(df$basin))
legend("topleft",
       legend = levels(factor(df$basin)),
       pch = 19,
       col = factor(levels(factor(df$basin))))
```
Looking at the data is possible to notice that the 2 points for the Pacific Tropical Cyclone are in the same cloud of points as the other data points for the Atlantic Tropical Cyclone given the same wind speed at landfall. Therefore the main situation/issue with the column "damage" is the lack of data, specially for Eastern Pacific Cyclones. 

```{r message = FALSE}
#install.packages("car")
library(car)
scatterplot(df$damage ~ df$wind_speed_landfall | df$basin, data = df, 
            smoother = TRUE, grid = TRUE, frame = FALSE)
```

5.  To understand the potential impact of climate change on tropical cyclone activity it may be useful to look at the evolution of a few metrics over the past 40 years. Answer the following questions based only on the data we have provided.

i.  Has the annual number of tropical cyclones making landfall been changing over the past 40 years? There are 3 possibilities: the annual number is increasing, there is no significant change, or it is decreasing. Identify which possibility is best supported by the data and explain why.

```{r}
df_grouped_year <-  group_by(df, year)
events <- summarize(df_grouped_year, number = n())
events
```

```{r}
model1 = lm(events$number ~ events$year, data=events)
plot(events$year, events$number, pch = 19)
legend("topleft",legend = paste("R2 is", format(summary(model1)$r.squared,digits=3)))
abline(model1,col='red') 
```
```{r}
summary(model1)
```

We can see that the number of report events has increased slightly over the years. The data might be telling us that there is an small increase of the number of events over the years.

ii. Has the average maximum sustained wind speed at landfall (that is, annual average of wind_speed_landfall) of tropical cyclones been changing over the past 40 years? There are 3 possibilities: the average maximum sustained wind speed at landfall is increasing, there is no significant change, or it is decreasing. Identify which of possibility is best supported by the data and explain why.

```{r}
df_grouped_year <-  group_by(df, year)
avg_speed <- summarize(df_grouped_year, avg = mean(wind_speed_landfall))
avg_speed
```

```{r}
model = lm(avg_speed$avg ~ avg_speed$year, data=avg_speed)
plot(avg_speed$year, avg_speed$avg, pch = 19)
legend("topleft",legend = paste("R2 is", format(summary(model)$r.squared,digits=3)))
abline(model,col='green') 
```
Based on the AVG Speed aggregated per year it is not possible to see a clear increase or decrease, specially as the R2 value is so low that it is safe to assume that the data is not correlated to the years. Therefore it seems that the speed of the events has changed with the years.

```{r}
model2 = lm(df$wind_speed_landfall ~ df$year, data=df)
plot(df$year, df$wind_speed_landfall, pch = 19)
legend("topleft",legend = paste("R2 is", format(summary(model2)$r.squared,digits=3)))
abline(model2,col='blue') 
```
After looking at the wind speed at landfill and the lack of correlation with the number of years, we can conclude than even though the number of events have increased by year, the strength of these tropical cyclones has not increased significantly.